The Eclipse Integrated Computational Environment

### 2018-07-12 src/review-response-20180712.tex

Dear Dr. Sobie,

Thank you very much for providing the reviewer's responses. Please find below a list of the issues that we have addressed and comments on the reviewer's responses. We found these reviews to be highly constructive and made every effort to address the reviews as thoroughly as possible.

We have resubmitted the manuscript and hope that you will reconsider it for publication in Software X.

Best,
Jay Jay Billings
ORNL

-----
Reviewer #1

>> It seems that comparison to other models could be improved and the authors could have listed features that provided by ICE that are either missing in other workflow models or are efficient compare, other models. Perhaps this section can be re-named as background or related work. Also, it was a little surprise that authors did not mention provenance which is one of the import features expected or desired by the scientific community.

We have expanded this section and addressed unique features and the comparison of ICE to other workflow engines with activities-based workflow models.

>> 
>> To summarize, the review,  the work seems interesting and useful to the community. The writing could be improved, below is the list of suggestions:
>> 
>> 1) A figure of some sort on complete architecture is extremely desired.

We have added four pictures and expanded the software architecture section to better describe ICE's architecture.

>> 
>> 2) Some of the references need to be looked up carefully. For instance reference 23 and 24 seem incomplete and less useful for a reader. See below:
>> 
>> [24] Jay Jay Billings. Eclipse ICE, October 2016.
>>

Five erroneous web citations were addressed.

>> 3) Renaming
>> Comparison to other models seems incomplete. Perhaps it could be renamed to something better such as "related work."
>>

Done.

>> 4) It is not very clear exactly how ICE integrate code from other languages. A reference to EASE was given but no other detail was provided.

We have added text to clarify that this is done by simple scripts in Javascript and Python through a built in shell, through workflow items, and in extreme cases through the Java Native Interface.

>> -----
>> Reviewer #2: This manuscript describes the Eclipse Integrated Computational Environment (ICE, https://www.eclipse.org/ice/), a scientific workbench and workflow environment for the execution and analysis of computational simulations. ICE adopts an alternative approach to typical workflow systems, which consists of defining analysis and computational processes into activities, as well as the support for workflow steering. Overall, the system has a proof community and relies in a well-established IDE (Eclipse), which increases its adoption by the community. This manuscript provides general concepts of the ICE system and illustrative examples, however lacks detailed information about the system architecture and execution environments. Additionally, related work is not comprehensive.

>> Detailed comments:
>> - Related work should be carefully revised, and include popular workflow system such as Swift, Pegasus, Makeflow, Taverna, Askalon, Nextflow, etc. (only Kepler is shown as an example). The manuscript claims that ICE follows the five categories described in the motivation section, however it is not the only workflow system that follows this classification.

The purpose of this work is not to extensively compare ICE to other workflow engines,

We do not believe that a detailed, n-by-n comparison of workflow systems is valuable in the context of a software publication on ICE, but we agree that it is important to illustrate the differences a little better. To that end, the text now discusses Chiron, Taverna, Aiida, and Fireworks. We have also added a reference to other work by the authors that describes the differences between modeling and simulation workflows and grid workflows in more detail.

>> The same issue appears again in Section 2.0.2 (section number should be revised too), where the presented model is claimed to significantly differ from similar efforts in workflow science. This is not accurate, there are several workflow systems that follows the activities model, such as Moteur, Taverna, and Chiron. It is crucial to distinguish ICE's contributions from these systems.

Taverna and Chiron are now discussed in the related work section. The exact differences between their Activities and ICE's are presented.

>> Last, interactive analysis of workflow executions is claimed as a novelty of this work; however workflow steering has been investigated by over a decade and support in systems such as Chiron and SciCumulus. How ICE differs from them?

Our paper only points out that ICE can perform this work and does not, as far as we can tell, claim that ICE uniquely provides this capability.

>> 
>> - Section 3, software description needs to be substantially improved. It misses an overview architecture (a figure would be extremely helpful) that shows how ICE components interact/leverage Eclipse components. This is one of the main contributions of this paper and should be better articulated.

We have addressed this point by adding new diagrams and more text describing the architecture of ICE. We have also added more information on the relationship to Eclipse Equinox, RCP, and PTP, etc. and cleaned up the sections that describe Eclipse across the document.

>> Several definitions are looseness:
>> + How the Items model differs from Taverna's model with Web Services?

Taverna's web services are now discussed specifically in the related work section, but since they are not similar to ICE's Items the relationship is not addressed. Shim services are now discussed briefly as well.

>> + There is a mention that ICE API will be evolved to make the API language closer to other systems. Which systems and how?

This is a generic statement about the names of classes and methods in other systems compared to the language used to describe these in ICE. For example, Item's in ICE are typically called tasks in Kepler-based systems such as Triquetrum.

The footnote has been changed to reflect this by adding "such as Triquetrum" after "other system" to provide an example.

>> + What defines a small output (so it will be automatically transferred)?

50MB. This has been updated in the text.

>> + Why is it more sophisticated? What can be achieved here? It is important to enumerate the advantages.

We have updated the text in this section to more clearly reflect our thinking on language-based workflows in ICE.

>> + What is a 'stubs' of plugins? How will it then interact with the framework? Which capabilities are used to enable them with ICE?

Updated the text to define stubs, strengthen the related statement on installation, and to describe the role of the OSGi in this process.

>> + Which operations are available via the APIs? It is mentioned that it has a rich API, but no detailed information is provided.

Detailed information is provided in the API documentation, which is cited in the final section. However, we strengthened this sentence by adding text that all workflow management functions are exposed by the API.

>> 
>> - IMHO, one of the main contributions and novelty of the paper is the combination of the Eclipse IDE with the workflow development/execution capabilities. However, it is not discussed in detail how a user may benefit from this integration.

This point has been addressed by creating a section specifically on the relationship to Eclipse out of content from the old framework extension section and new information on the benefits of self-hosting.

>> 
>> - The examples section would significantly benefit of more information about how the workflows are executed.

All workflows in ICE are executed as described in the executing jobs sub section.

>> 
>> - Also, there are mentions that ICE can run locally and in HPC platforms, but no information is provided on how the system actually executes the workflows. There is a quick mention to SSH connections, but nowadays there are other methods to remotely submit jobs for execution, such as Globus GRAM, Bosco, etc. The manuscript needs a proper workflow execution section which describes the different methods how the workflows can be executed.

The executing jobs section completely covers how workflows are executed in ICE. We have modified it to state that Globus GRAM, etc. are not supported. We have also provided some text to clarify that executing jobs is not the same as executing a workflow.

>> 
>> In summary, I see much potential in the proposed workflow tool, however a thoroughly characterization and description of the architectural components and their capabilities is key to demonstrate the novelty and impact of the discussed system.

We are very grateful for this thorough review of the paper. We believe that we have addressed this review just as thoroughly with the significant updates provided in this revised manuscript.

Also tightened up text significantly.

>> 
>> ******************************************
>> Please note that the editorial process varies considerably from journal to journal. To view the submission-to-publication lifecycle, click here: http://help.elsevier.com/app/answers/detail/p/7923/a_id/160
>> 
>> For further assistance, please visit our customer support site at http://help.elsevier.com/app/answers/list/p/7923. Here you can search for solutions on a range of topics, find answers to frequently asked questions and learn more about EES via interactive tutorials. You will also find our 24/7 support contact details should you need any further assistance from one of our customer support representatives.

### 2018-07-12 src/content.tex
Addressed all reviews and finished grammatical edits. Signed-off-by: Jay Jay Billings <billingsjj@ornl.gov>

It is not clear which, if any, of these solutions is better than the
others, and practical requirements will ultimately dictate the path of a
project's progress. This work considers a middle ground solution and
presents the Eclipse Integrated Computational Environment (ICE) as proof
that it is possible to create such a system. Specifically, the work
described here shows that:

@@

\subsubsection{Workflow Model}\label{workflow-model}

@@

In this work, it should be assumed---unless otherwise noted---that
``executing a job'' includes monitoring that job in one or more ways,
possibly including real-time updates to visualizations.

\textbf{Analyzing results} includes executing special jobs to transform
data in one or more prescribed ways and producing artifacts with
scientific significance from the transformed data. This could include, for
example, postprocessing results and visualizing the new data with
dedicated visualization tools. For many types of scientific computing,
this includes viewing the results of a simulation on a mesh or grid and
extracting publication-quality images or movies from that data. Other
cases might include analyzing results in preparation for follow-on
simulations or performing feature extraction, classification, or
activities for machine learning and data mining.

@@

\textbf{Modifying code} is not typically considered a part of a
scientific computing workflow. However, modeling and simulation use
cases often require users to explicitly modify code before execution or to issue special re-build instructions. This is true, for example, with the computational fluid dynamics code Nek5000 which requires modification of code before compilation and special build instructions using the \textit{makenek} shell script,
\cite{the_nek5000_team_nek5000_2014},
Many computational scientists consider ``their workflow'' to be re-running
software after modifications for purely exploratory purposes. This might
be required if the model that the author is modifying cannot be
configured directly as part of the input but can be easily accomplished by manipulating the source code.

\subsubsection{Comparison to Other
Models}\label{comparison-to-other-models}

This model of workflows differs significantly from those of similar
efforts in workflow science because it defines workflows in terms of
activities. Other workflow models in the literature define a workflow as
a collection of computing processes. For example, Yu and Buyya define
grid workflows as ``a collection of tasks that are processed on
distributed resources in a well-defined order to accomplish a specific
goal,'' \cite{yu_taxonomy_2005}. Others, such as Pizzi et al., subscribe to
similar definitions \cite{pizzi_aiida:_2016}. This ``process'' view is
acceptable where the workflow is static and does not require additional
human input or ``human in the loop'' behavior after all the initial
human input is provided. However, workflows within ICE are fully interactive with regular callbacks to humans. It is simpler to discuss ``activities'' than it is to create a
distinction between ``human processes'' and ``computer processes.''
Focusing on ``activities'' over processes (human or computer) also has
the benefit of removing concrete elements such as hardware or software properties that distract from details of workflows and workflow 
management systems. That is, considerations such as memory usage and 
raw performance are important, but questions about the abstract workflow 
or what the workflow management system should do are far more important in this context.

@@

ICE was specifically created to address hands-on workflows for computational scientists, as in section \ref{illustrative-examples}. Users download and
execute ICE locally, and ICE in turn orchestrates local or remote workflows as
required. It provides a comprehensive workbench for modeling and
simulation that includes tools for workflows, visualization, data
management, and software development.

@@

Workflows and tasks in ICE are not explicitly treated as trees, or
directed acyclic graphs (DAGS), as is common with grid workflow tools
\cite{yu_taxonomy_2005}. Instead, ICE's design is inspired by representational
state transfer (REST) \cite{fielding_architectural_2000}.

Users initiate requests to create, edit, update, or delete workflows from
the ``ICE Client'' (the workbench). The list of available workflows
that can be created is provided dynamically to the ICE Client by
the ``ICE Core,'' which acts as a server. This information is
provided dynamically because it often changes at run time based on 
both the configuration of available workflow components in the registry and on
persisted workflows that users have saved in their workspace.
Information about workflows is provided to the Client by the Core
through stateless ``Forms'' that describe the workflow and provide
all the necessary information to understand \emph{what} should process
the workflow (but not \emph{how} it should be processed). Once users
modify the description of the workflow in the Form to provide their
specific details, the Client dispatches a request to the Core to modify
and/or process the workflow. The Core then uses the information from the Form
to perform a service lookup to identify the correct service to process the workflow.

Workflows in ICE are encoded in and processed by services called ``Items,'' and each
workflow type is a subclass of an Item base class. Items are registered dynamically
through a service registry in ICE (see Section \ref{framework}) and
provide the Forms for the ICE Core and Client. Since ICE's design is
highly object oriented, it is easiest to think of the Item class as a
description of an abstract workflow and an Item object (an instance of
the class) as a concrete workflow with all required execution details
provided by its Form.

Individual components of workflows (i.e., workflow ``tasks'' or
``nodes'') are either encoded directly in the workflow's subclass of
Item or provided as ``Actions'' that are dynamically registered with
an ``Action Factory'' and obtained at run time. Table 1 describes the
differences between Items, Actions, and Forms.\footnote{An upcoming 
update to the API will include the formal introduction of
IWorkflow, IWorkflowTask, and IWorkflowEngine interfaces to bring ICE's
API language closer to other systems. However, this must be done carefully to
preserve backwards compatibility with the present API.}

@@

In addition to running as a desktop workbench, ICE can be run as a
headless web server with a remote service interface and a web
API. The web API is also used as the primary means of providing real-time 
feedback and monitoring support in ICE.

\subsubsection{Item States}\label{item-states}

@@

This design is very important. First, it means that all workflows in
ICE, regardless of their actual details and functions, can only behave in
a specific set of known and predictable ways, and this predictability 
simplifies the way the Core manages and interacts with Items. Second, by
formalizing state and error checks, ICE explicitly delineates which workflows 
can be executed from those that must receive additional configuration. 
Finally, it allows developers implementing Items and
Actions to specify, by contract, what is required before proceeding to the
next task, processing the workflow, or declaring a successful execution.

@@

When workflows are created and modified, ICE saves permanent copies of 
Forms to disk in a special directory called a ``workspace.''
Workspaces can contain projects, folders, and files, including data,
code, input, and output. ICE automatically manages local and remote (or
even local \emph{to} remote) transfers of data files when executing
workflows if the files are detected in the same directory of the
workspace as the workflow itself. For example, when executing a remote
job, ICE will automatically move the input file if it is specified in
the workflow and available in the project directory of the workspace.
Likewise, if the output is small enough, ICE will automatically move it back
to the local directory. By convention, all paths in ICE's Forms are
relative to the workspace root path. Workspace directories are specified
by the user and are fully functional Eclipse workspaces.

@@

\subsubsection{Extending the Framework to Add Custom
Workflows}\label{framework}

ICE is an Eclipse Rich Client Platform (RCP) application
\cite{mcaffer_eclipse_2010} and has a plugin architecture based on 
Equinox, which is the reference implementation of the Open Service Gateway Initiative (OSGi) 
framework specification \cite{mcaffer_osgi_2010}. ICE uses more than 1,200 additional
packages from the Eclipse ecosystem to provide services like language
support and visualization. Each unique element of ICE described
in this work---including the Client, Core, Items, and data structures---are 
provided as plugins to Equinox. Most plugins are managed
dynamically and provided as services that can be obtained through a
service registry. Most file I/O in ICE, with only a few exceptions,
interacts with the RCP's resources plugin. This also includes remote
resources that are managed with the Eclipse Parallel Tools Platform
\cite{tibbitts_integrated_2009}.

It is not possible to create new Items in ICE graphically, and---unless 
they provide their own plugin to the framework---users can
use only the workflow Items provided by their version of ICE. This extension method can seem like an onerous task to the novice user, especially if they do not know Java, but
this design was chosen because it is, in the opinion of the authors, a
far more sophisticated and sustainable way to create workflows than, for example,
using a wiring diagram.

ICE provides built-in development tools to automatically generate empty but ready to compile classes, (``stubs''), of
plugins that can then be installed into the framework as services. This
``self-hosting'' makes it possible for new users to create complex,
sophisticated workflows very quickly because they are not required to
know how to interact with the framework. Furthermore, ICE provides a
rich API that that exposes all workflow management functionality, documentation, tutorials, and tools to further simplify the process.

There are many situations where configuring workflows graphically is
unacceptable, such as when a very large number of workflows will be
executed or the type of information required is very fine grained. In
these cases, it is often necessary to provide scripts to the workflow
management system. ICE includes the Eclipse Advanced Scripting
Environment (EASE)~\cite{pontesegger_eclipse_2015} for scripting because it
provides a way to script Eclipse RCP projects natively in Javascript,
Jython, and Python. This also makes it possible to extend the
environment by adding Items in these languages.

@@

Outside of the development team, ICE is commonly deployed as a
sophisticated user environment for computational science projects 
(see Section \ref{impact}) and as a visualization tool. The ICE source code originally contained a
significant amount of visualization support, but at the request of users
in the community that support was ``spun off'' in early 2016 as the 
Eclipse Advanced Visualization Project (EAVP)~\cite{billings_eclipse_2015}.

@@

Pannala et al. developed a Virtual Integrated Battery Environment (VIBE)
as part of their research into safety and performance characteristics of
lithium ion batteries \cite{pannala_multiscale_2015}. VIBE includes ICE
as part of its distribution, and new workflows were added to ICE to enable users
to add multiple types of input, configure the simulation software, and launch
simulations of virtual batteries. Interactive 3D visualizations of the
results were embedded in the launcher so that users can quickly find
their results. Figure 1 shows a (simulated) prismatic-cell battery's temperature
distribution during discharge.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{images/vibe_20151016.png}
\caption{ICE workbench for VIBE analysis.}
\end{figure}

@@

The MOOSE Framework is a powerful, easy-to-use multiphysics framework
developed at Idaho National Laboratory \cite{gaston_moose:_2009}. ICE
provides workflow tools for MOOSE as well as specialized class
generation utilities for developing custom MOOSE kernels. Many of the
MOOSE tools in ICE were developed closely with the MOOSE team to
reproduce various aspects of MOOSE's user interface, known as ``Peacock.'' 
Figure 2 shows an example of the ICE workbench for a simple structural mechanics
problem solved using the MOOSE framework \cite{mccaskey_scientific_2015}.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{images/ice-moose.png}
\caption{ICE workbench for MOOSE workflows.}
\end{figure}

@@

Solid-state sintering of parts printed using binder jetting
significantly increases part strength by decreasing the part porosity
and eliminating voids. However, this process does cause the resulting
product to shrink and warp from its original layout. An ideal, near-net-shape, 
process would combine binder jetting with solid-state sintering and accounts
for warpage and deformation in the part's design phase. 
Figure 3 shows an ICE-based workbench for performing simulations of
this process with visualizations of the pre- and post-simulation
properties of a central body with eight cantilevers. The primary
deformation in this type of geometry is bending or drooping of the
cantilevers due to sintering and thermal creep.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{images/ice-bjm.png}
\caption{ICE workbench for binder jet modeling.}
\end{figure}

>>>>

It is not clear which, if any, of these solutions is better than the
others, and practical requirements will ultimately dictate the path of a
project's progress. This work considers a middle ground solution and
presents the Eclipse Integrated Computational Environment (ICE) as proof
that it is possible to create such a system. Specifically, the work
described here shows that

@@

\subsection{Workflow Model}\label{workflow-model}

@@

Local jobs in ICE are executed using standard Java system calls. Remote jobs are launched only through SSH connections on remote machines. This includes direct SSH command execution on clusters and proxy connections through a pilot service on large Leadership-class supercomputers. Services such as Globus GRAM and Bosco are not supported, but the SSH command execution includes extensive support for numerous batch and queuing systems. The Eclipse Parallel Tools Platform (PTP) is used to create all remote SSH connections, regardless of the target machine \cite{tibbitts_integrated_2009}.

In this work, it should be assumed---unless otherwise noted---that
``executing a job'' includes monitoring that job in one or more ways,
possibly including real-time updates to visualizations. It is also important to note that executing a job is not the same as executing a workflow. Executing jobs specifically refers to launching simulations, whereas executing a workflow could be something different such as generating input or post-processing results. This is an important distinction because ICE's workflows are all executed locally, but simulations and work may be distributed remotely depending on the implementation of the workflow plugin.

\textbf{Analyzing results} includes executing special jobs to transform
data in one or more prescribed ways and producing artifacts with
scientific significance from the transformed data. This could include, for
example, post-processing results and visualizing the new data with
dedicated visualization tools. For many types of scientific computing,
this includes viewing the results of a simulation on a mesh or grid and
extracting publication-quality images or movies from that data. Other
cases might include analyzing results in preparation for follow-on
simulations or performing feature extraction, classification, or
activities for machine learning and data mining.

@@

\textbf{Modifying code} is not typically considered a part of a
scientific computing workflow. However, modeling and simulation use
cases often require users to explicitly modify code before execution or to issue special re-build instructions. This is true, for example, with the computational fluid dynamics code Nek5000, which requires modification of code before compilation and special build instructions using the \textit{makenek} shell script
\cite{the_nek5000_team_nek5000_2014}.
Many computational scientists consider ``their workflow'' to be re-running
software after modifications for purely exploratory purposes. This might
be required if the model that the author is modifying cannot be
configured directly as part of the input but can be easily accomplished by manipulating the source code.

\subsection{Related Work}\label{comparison-to-other-models}

ICE's model of workflows differs significantly from many other
efforts in workflow science because it defines workflows in terms of high-level activities that are meant to guide workflow implementers and users. Many other workflow models in the literature define a workflow as
a collection of computing processes. For example, Yu and Buyya define
grid workflows as ``a collection of tasks that are processed on
distributed resources in a well-defined order to accomplish a specific
goal'' \cite{yu_taxonomy_2005}. Others, such as Pizzi et al., subscribe to
similar definitions \cite{pizzi_aiida:_2016}. This ``process'' view is
acceptable where the workflow is static and does not require additional
human input or ``human in the loop'' behavior after all the initial
human input is provided. However, workflows within ICE are fully interactive with regular callbacks to humans. It is simpler to discuss ``activities'' than it is to create a distinction between ``human processes'' and ``computer processes.''
Focusing on ``activities'' over processes (human or computer) also has
the benefit of removing concrete elements such as hardware or software properties that distract from details of workflows and workflow 
management systems. That is, considerations such as memory usage and 
raw performance are important, but questions about the abstract workflow 
or what the workflow management system should do are far more important in this context.

There are other workflow engines that have their own concepts of activities, with various similarities and differences compared to ICE's model. For example, Taverna has a concept of activities \cite{wolstencroft_taverna_2013}. Activities in Taverna are defined as invocable activities within a workflow, which is closely related to a lower-level class in ICE called Actions (see \ref{item}). On the other hand, ICE's activities loosely describe what can be done conceptually and the workflows that are implemented provide the Actions that carry out these activities.

ICE and Taverna can both natively execute Java code, but ICE 2.0 cannot natively invoke Web Service Definition Language (WSDL) web services. ICE uses regular Java-based libraries to interact with web services, such as Jersey, and requires that developers write the code to do this directly. Shim services in Taverna are provided in ICE as separate services in the framework, but are otherwise very similar.

Chiron, an algebraic workflow engine, has a concept of activities that is closer to ICE's than Taverna's, but still very strongly related to implementation \cite{ogasawara_chiron_2014}. Activities in Chiron are combinations of the program or expression to execute, plus the input schema, plus the output schema. This is a close match to the design of the Item class in ICE. ICE's Items (see Section \ref{item}) implement the high-level activities, such as launching a job or creating a model. The major difference between Chiron's activities and ICE's Items appears to be scope. A Chiron activity such as Map or Reduce would normally be implemented as an Action in ICE, whereas an Item in ICE may be tasked with executing multiple Actions to enact a workflow.

One significant difference between ICE and many other systems is that it uses a language-based approach to defining workflows, but with a standard language, Java, as opposed to a custom workflow definition language. Pizzi et al. do this with Aiida and Python, and the Fireworks Workflow Engine also uses Python. The advantage of this approach is that compilation is not a problem. ICE also differs from other workflow engines because it is specifically focused on modeling and simulation instead of grid workflows. Grid workflows are almost always defined as workflows based on trees without humans in the loop, but modeling and simulation workflows are much more interactive and undirected \cite{billings_toward_2017}.

@@

ICE was specifically created to address hands-on workflows for computational scientists, as in Section \ref{illustrative-examples}. Users download and
execute ICE locally, and ICE in turn orchestrates local or remote workflows as
required. It provides a comprehensive workbench for modeling and
simulation that includes tools for workflows, visualization, data
management, and software development.

@@

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{src/images/ICE-arch-1.jpg}
\caption{High-level architecture of ICE showing the relationships and rough division of responsibility between the five primary components of ICE. The ICE Core is responsible for the workflow management and delegations instructions and data to the Item and Data Structures, respectively. User input is received from the client or through headless interactions, while persistence of large amounts of data is delegated to a persistence service.}
\label{highlevel-arch}
\end{figure}

Workflows and tasks in ICE are not explicitly treated as trees, or
directed acyclic graphs (DAGS), as is common with grid workflow tools
\cite{yu_taxonomy_2005}. Instead, ICE's design is inspired by representational
state transfer (REST) and the workflow engine is implemented as services in an Eclipse Rich Client Platform (RCP) application
\cite{fielding_architectural_2000} \cite{mcaffer_eclipse_2010}.

Figure \ref{highlevel-arch} shows the five primary components of ICE and their relationships to each other. Users initiate requests to create, edit, update, or delete workflows from the ``ICE Client'' (the workbench) or through headless interactions using web or language programming interfaces. The list of available workflows
that can be created is provided dynamically to the ICE Client by
the ``ICE Core,'' which acts as a server and is the primary component responsible for workflow orchestration. Workflow information is
provided dynamically because it often changes at run time based on 
both the configuration of available workflow components in the registry and on
persisted workflows that users have saved in their workspace. Workflow task enactment and execution is performed by services called ``Items.'' Information about workflows is provided to the Client by the Core
through common data structures rooted in a stateless ``Form'' tree. Forms describe the workflow and provide
all the necessary information to understand \emph{what} should process
the workflow (but not \emph{how} it should be processed). Once users
modify the description of the workflow in the Form to provide their
specific details, the Client dispatches a request to the Core to modify
and/or process the workflow. The Core then uses the information from the Form
to perform a service lookup to identify the correct Item to process the workflow.

\subsubsection{ICE Core}

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{src/images/ICE-arch-2.jpg}
\caption{The architecture of the ICE Core and its dependencies. The ICE Core class handles communications between workflow Items and common data structures, but delegates construction of workflow Items to builders.}
\label{core-arch}
\end{figure}

The primary purpose of the ICE Core is to orchestrate workflows and services for the platform. Figure \ref{core-arch} shows the primary classes that interact with the ICE Core. It is the ``core'' control component of ICE where all other component interactions converge. It is the middle tier of ICE's architecture and brokers communications between Items and Clients (graphical or not) by sharing Forms that contain workflow data. The ICE Core class delegates construction and management of Items to the Item Manager.

The Item Manager acts as a container and factory for Items created by the ICE Core. Items are constructed using the Builder pattern, with concrete construction handled by realizations of the ItemBuilder interface.

ICE 2.0 includes a graphical client based on the Eclipse Workbench, which is common for Eclipse applications. The workflow management capabilities of the core can be accessed through languages and a web-based programming interface as well (see Section \ref{framework}).

\subsubsection{Items}
\label{item}
\label{item-states}

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{src/images/ICE-arch-3.jpg}
\caption{The class hierarchy of ICE Items showing the relationships between the Item, its subclasses, and services that it uses. Vibe Launcher and MOOSE Launcher are examples of special Job Launch Items for executing jobs, and their applications are discussed in section \ref{illustrative-examples}.}
\label{item-arch}
\end{figure}

Each workflow Item in ICE is an independent service, and each
workflow type is a subclass of the Item base class or another Item subclass. Items are registered dynamically
through a service registry in ICE (see Section \ref{framework}) and
provide collections of workflow data, Forms, for the ICE Core and Client. Since ICE's design is
highly object oriented, it is easiest to think of the Item class as a
description of an abstract workflow and an Item object (an instance of
the class) as a concrete workflow with all required execution details
specified in its Form.

Figure \ref{item-arch} shows the class hierarchy of the Item class and its collaborators.
Individual components of workflows (i.e., workflow ``tasks'' or
``nodes'') are either encoded directly in the workflow's subclass of
Item or provided as ``Actions'' that are dynamically registered with
an ``Action Factory'' and obtained at run time. Common data structures that are shared across Items, the Client, and the Core are provided by the data structures component. Table 1 describes the differences between Items, Actions, and Forms.\footnote{An upcoming 
update to the API will include the formal introduction of
IWorkflow, IWorkflowTask, and IWorkflowEngine interfaces to bring ICE's
API language closer to other systems such as Triquetrum. However, this must be done carefully to preserve backwards compatibility.}

@@

Commonly used subclasses of Item include the Model and Job Launcher classes, which can be  used to create workflow plugins for input generation and workflow execution, respectively. Composite Items are special Items that are built of compositions of other Items. Composite Items are commonly used for workflow Items that can be executed in parallel or to tightly couple different workflow tasks for resource and performance management.

@@

This design is very important. First, it means that all workflows in
ICE, regardless of their actual details and functions, can only behave in
a specific set of known and predictable ways, and this predictability 
simplifies the way the Core manages and interacts with Items. Second, by
formalizing state and error checks, ICE explicitly delineates which workflows 
can be executed from those that must receive additional configuration. 
Finally, it allows developers implementing Items and
Actions to specify by contract what is required before proceeding to the
next task, processing the workflow, or declaring a successful execution.

\subsubsection{Data Structures}

ICE contains a collection of common data structures that are useful for representing data needed modeling and simulation workflows. Figure \ref{data-arch} shows some of these classes and their relationship. All data structures (and Items) in ICE are uniquely identifiable and behave in predictable ways: they receive and dispatch updates, can be visited to determine type information, etc. \textit{Components} in the data sense, not the class sense, represent a special data structure that forms a child of a Form. Forms themselves are components and hierarchically aggregate other components.

The name ``Component'' in the context of ICE's data structures refers to a unique and reusable collection of data, not a class or component in the object oriented sense. Components that hold data do very little of their own work but strongly scope the interfacial contract of their clients.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{src/images/ICE-arch-4.jpg}
\caption{Common data structures and their relationships in ICE.}
\label{data-arch}
\end{figure}

@@

When workflows are created and modified, ICE saves permanent copies of 
Forms to disk in a special directory called a ``workspace.''
Workspaces can contain projects, folders, and files, including data,
code, input, and output. ICE automatically manages local and remote (or
even local \emph{to} remote) transfers of data files when executing
workflows if the files are detected in the same directory of the
workspace as the workflow itself. For example, when executing a remote
job, ICE will automatically move the input file if it is specified in
the workflow and available in the project directory of the workspace.
Likewise, if the output is small enough (less than 50 megabytes), ICE will automatically move it back
to the local directory. By convention, all paths in ICE's Forms are
relative to the workspace root path. Workspace directories are specified
by the user and are fully functional Eclipse workspaces.

@@

\subsubsection{Relationship to Other Eclipse Technologies}\label{framework}

ICE is an Eclipse Rich Client Platform (RCP) application
\cite{mcaffer_eclipse_2010} and has a plugin architecture based on 
Equinox, which is the reference implementation of the Open Service Gateway Initiative (OSGi)
framework specification \cite{mcaffer_osgi_2010}. It is most appropriate to think of Eclipse ICE as an alternative flavor of any other Eclipse development environment because like those environments it is just a collection of Eclipse plugins with a well-defined purpose and brand. ICE uses more than 1,200 additional packages from the Eclipse ecosystem to provide services like language
support and visualization. Each unique element of ICE described
in this work---including the Client, Core, Items, and data structures---is 
provided as plugins to Equinox. Most plugins are managed
dynamically and provided as services that can be obtained as dynamic OSGi Declarative Services. File I/O in ICE, with only a few exceptions,
interacts with the RCP's Resources plugin and the standard Eclipse Workspace. This also includes remote resources that are managed with Eclipse PTP.

ICE plugins are created used normal Eclipse development tools and installed into either the running instance of ICE or published in a newly built version. In either case, the running platform of ICE can only use the plugins that are actually installed and at least a small amount of code must be written in Java for each plugin. This extension method may seem odd to the novice user or developer, especially if they do not know Java very well, but
this design was chosen because it is an easy, fast, and precise way to create workflows.

ICE provides built-in development tools to automatically generate empty but ready to compile classes (``stubs'') of plugins that can then be installed into the framework as services. These tools are based on other code generation tools in the Eclipse platform. This
``self-hosting'' makes it possible for new users to create complex,
sophisticated workflows very quickly because they are not required to
know how to interact with the framework. Development of new workflows is streamlined because ICE provides a
rich API that exposes all workflow management functionality, documentation, tutorials, and tools to further simplify the process. These are shared through the self-hosted environment by code completion tools, help menus, and documentation overlays. The close proximity of the workflow environment with the development platform for modeling and simulation codes is also beneficial because, as a self-hosted entity, workflows can be co-developed with the code.

ICE can be run as a headless web server with a remote service interface and a web
API. The web API is also used as the primary means of providing real-time 
feedback and monitoring support in ICE, and it is published as an OSGi service as well. These services can also be consumed as OSGi Remote Services.

There are many situations where configuring workflows graphically or serially is
unacceptable, such as when a very large number of workflows will be
executed or the type of information required is very fine grained. In
these cases, it is often necessary to provide scripts to the workflow
management system. ICE includes the Eclipse Advanced Scripting
Environment (EASE)~\cite{pontesegger_eclipse_2015} for scripting because it
provides a way to script Eclipse RCP projects natively in Javascript,
Jython, and Python. This also makes it possible to extend the
environment by adding Items in these languages. Specifically, scripts in Javascript and Python are written and executed in simple shell that is part of the workbench. Calling other languages, such as C or Fortran, is typically accomplished by adding workflow Items for the executables in these languages and performing local workflow job launches. C languages routines have been called from ICE through the Java Native Interface (JNI), but this is not a technique that is widely used with the platform because of its complexity.

@@

Outside of the development team, ICE is commonly deployed as a
sophisticated user environment for computational science projects 
(see Section \ref{impact}) and as a visualization tool. The ICE source code originally contained a
significant amount of visualization support, but at the request of users
in the community, that support was ``spun off'' in early 2016 as the 
Eclipse Advanced Visualization Project (EAVP)~\cite{billings_eclipse_2015}.

@@

Pannala et al. developed a Virtual Integrated Battery Environment (VIBE)
as part of their research into safety and performance characteristics of
lithium ion batteries \cite{pannala_multiscale_2015}. VIBE includes ICE
as part of its distribution, and new workflows were added to ICE to enable users
to add multiple types of input, configure the simulation software, and launch
simulations of virtual batteries. Interactive 3D visualizations of the
results were embedded in the launcher so that users can quickly find
their results. Figure \ref{vibe} shows a (simulated) prismatic-cell battery's temperature
distribution during discharge.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{images/vibe_20151016.png}
\caption{ICE workbench for VIBE analysis.}
\label{vibe}
\end{figure}

@@

The MOOSE Framework is a powerful, easy-to-use multiphysics framework
developed at Idaho National Laboratory \cite{gaston_moose:_2009}. ICE
provides workflow tools for MOOSE as well as specialized class
generation utilities for developing custom MOOSE kernels. Many of the
MOOSE tools in ICE were developed closely with the MOOSE team to
reproduce various aspects of MOOSE's user interface, known as ``Peacock.'' 
Figure \ref{moose} shows an example of the ICE workbench for a simple structural mechanics
problem solved using the MOOSE framework \cite{mccaskey_scientific_2015}.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{images/ice-moose.png}
\caption{ICE workbench for MOOSE workflows.}
\label{moose}
\end{figure}

@@

Solid-state sintering of parts printed using binder jetting
significantly increases part strength by decreasing the part porosity
and eliminating voids. However, this process does cause the resulting
product to shrink and warp from its original layout. An ideal near-net-shape 
process would combine binder jetting with solid-state sintering and account
for warpage and deformation in the part's design phase. 
Figure \ref{binder} shows an ICE-based workbench for performing simulations of
this process with visualizations of the pre- and post-simulation
properties of a central body with eight cantilevers. The primary
deformation in this type of geometry is bending or drooping of the
cantilevers due to sintering and thermal creep.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{images/ice-bjm.png}
\caption{ICE workbench for binder jet modeling.}
\label{binder}
\end{figure}
